The Anthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-06-02 paper.tex
integrate ontological architecture

\section{Ontological Architecture}\label{sec:architecture}
%Gesamtarchitektur, Zusammenhang zwischen GFO, ANNOdc, ANNOds, 3-Ontologien-Methode, alles mit Verweisen auf die nächsten Kapitel, vielleicht ein kleines Bild

%The development of ANNO and its integration into \aw{} follows the \emph{three ontology method}~\citep{threeontologymethod} for ontology-based software engineering:
%First, \emph{task ontology} is created as a conceptual model of \aw{}.
%Second, the General Formal Ontology (GFO)~\citep{gfo} is chosen as \emph{top-level ontology} to provide a theoretical foundation.
%Third, the \emph{domain ontology} is created that describes the domain of skeletal anatomy.
\todo{figure out how to best separate ontology and software}
The ontological architecture of ANNO consists of three interrelated layers.

\begin{enumerate}

 \item ANNOds (\cref{sec:domain}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating
the parts of human skeleton (bones, teeth and landmarks, such as, mandible, left mandibular central secondary incisor tooth
or mentale dexter) as well as to model their properties and relations (e.g., the distance between mentale dexter and mentale sinister)
required to derive human phenotypes (e.g., sex or height). The ANNOds is embedded in ANNOdc
(i.e., the ANNOds classes are subclasses of the ANNOdc classes).

 \item ANNOdc (\cref{sec:core}, \cref{fig:core}) is a domain-core (dc) ontology to provide the core entities of a domain, such as, inter alia,
the general anatomical categories (\aurl{anno}{Anatomical\_structure} and \aurl{anno}{Landmark}), categories for describing different characteristics of
landmarks (\aurl{anno}{Landmark\_location}, \aurl{anno}{Landmark\_property} and \aurl{anno}{Landmark\_line\_region}) as well as
the category \aurl{anno}{Phenotype} to model the rules for determining human phenotypes. The ANNOdc is founded by \emph{General Formal Ontology} (GFO)~\citep{gfo}.

 \item GFO is a top-level ontology and is used for integrating and founding the ANNO. In the current paper, we refer especially to the categories
\enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO (\cref{sec:core}, \cref{fig:gfo}).
\end{enumerate}

This architecture fits very well with the three-ontology method~\citep{threeontologymethod} on the basis of which we designed
the \aw{} software (\cref{sec:aw}) for annotating human skeletons and determining phenotypes.
This method for developing consistent and modular software is based on the interaction of a task ontology, a domain ontology
and a top-level ontology. In our case, the ANNOdc plays the role of a task ontology, while the ANNOds acts as a domain ontology.
One of the advantages of the three-ontology method is that the software only needs to implement the access to entities (classes, properties)
of the task ontology, whereas the entities of the corresponding domain ontology are processed dynamically.
The \aw{} uses the ANNOdc as an interface to access the ANNOds entities, e.g. by listing all subclasses of a specific ANNOdc class,
such as \aurl{anno}{Bone}, as described in \cref{sec:aw}.

>>>>

\section{Ontological Architecture}\label{sec:architecture}

The ontological architecture of ANNO consists of three interrelated layers representing by a top-level ontology, a domain-core ontology and a domain-specific ontology:

\begin{enumerate}
\item We use the General Formal Ontology (GFO)~\citep{gfo} as top-level ontology of ANNO.
In the current paper, we refer especially to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO (\cref{sec:core}, \cref{fig:gfo})

\item ANNOdc (\cref{sec:core}, \cref{fig:core}) is a domain-core (dc) ontology to provide the core entities of a domain.
This includes general anatomical categories (Bone and Tooth), a category for describing their characteristics (Anatomical property), anatomical spatial entities (Anatomical space, surface, line and point) as well as the category \emph{Phenotype} to model the rules for determining human phenotypes.
ANNOdc is embedded in the GFO, i.e. the ANNOdc classes are subclasses of GFO classes.

\item ANNOds (\cref{sec:domain}) is a domain-specific (ds) ontology for describing domain-specific entities to be used for annotating the parts of the human skeleton.
These are bones, teeth, their parts and compounds, such as \emph{mandible}, \emph{Mental protuberance} or the facial skeleton.
It is also used for modelling their properties and relations, such as the distance between Mentale dexterum and Mentale sinistrum, needed to derive human phenotypes like sex or height.
ANNOds is embedded in ANNOdc.
\end{enumerate}
